Comparing Exchangeable Effects (EE) and Exchangeable Standardized Effects (ET) models

### 2016-10-19 analysis/CompareEEvsETmodel.Rmd
Working on revisions to CompareEEvsETmodel.Rmd.

The ashr package implements a shrinkage-based Empirical Bayes
method for estimating the values of effects $\beta_j$ based
on estimates of the effects ($\hat\beta_j$) and their standard errors ($s_j$).

The default model is that the effects $\beta_j$ are identically distributed
from a unimodal distribution $g$. 
In particular, ashr assumes that $\beta_j$ is independent of the standard error $s_j$.
That is,
$$\beta_j | s_j \sim g(\cdot)     \quad (*).$$

Here we consider the more general model:
$$\beta_j/(s_j)^\alpha | s_j \sim g(\cdot)   \quad (**).$$
The case $\alpha=0$ is the model (*) above. The case $\alpha=1$ assumes that 
the *t statistics* $\betahat_j/s_j$ are identically distributed from a unimodal
distribution. Under this assumption the expected size of the *unstandardized* effect $\beta_j$ depends on the standard error $s_j$: the larger $s_j$ is, the larger (in absolute value) $\beta_j$ is expected to be.

By analogy with Wen and Stephens (AoAS), we refer to the model (*) as the EE model ("exchangeable effects"), and the model (**)
as the ET model ("exchangeable T statistics model")

What might motivate the ET model? 
We can provide two distinct motivations. First, suppose for concreteness that the effects
$\beta_j$ reflect differences in gene expression between two conditions. One factor that affects
$s_j$ is the variance of gene $j$ within each condition. One could imagine that, perhaps, genes with a larger variance within each condition are less tightly regulated, and therefore more likely to show a large difference between conditions (i.e. large $\beta_j$) than genes with a small variance. This provides
a biological motivation for the possibility that larger $s_j$ might 
correlate with larger $\beta_j$ (although of course not for the exact functional form above).
A second motivation is more statistical: it turns out that this assumption is in some sense
the implicit assumption made by existing methods to fdr analysis based on p values. 
More specifically, under this alternative assumption, when 
attempting to identify ``significant" effects, the empirical Bayes approach will rank the genes 
in the same way as the usual $p$ values computed from $\hat\beta_j/s_j$.

It is straightforward to use the ashr package to perform analysis under the ET model:
simply specify alpha=1. (Internally, this causes ash to 
replace betahat with the standardized betahat, $\hat\beta_j/s_j$, and the standard errors for these standarized betahat values with 1; there there is some bookkeeping to be done to make sure we return the right likelihoods and posteriors for the original beta, and not for these standardized values... ash takes care of this.) It is also straightforward to compare
the two competing modelling assumptions (EE vs ET) by computing the log likelihood ratio, $log[p_{EE}(\hat\beta | s, \hat g_{EE})/p_{ET}(\hat\beta | s, \hat g_{ET})]$.

We now illustrate by a simulated example. We assume that the standard errors come
from a gamma distribution, and then generate the effects $\beta_j$ under the ET model so that genes with bigger standard errors tend to have bigger effects.
```{r}
  set.seed(1234)
  nsamp=1000
  betahat.se = rgamma(nsamp,1,1)
  beta = betahat.se * rnorm(nsamp) #simulate effects under the alternative assumption 
  betahat = rnorm(nsamp,beta,betahat.se)
  zscore = betahat/betahat.se
  pval = pchisq(zscore^2,df=1,lower.tail=FALSE)
  plot(betahat, -log(pval))
```

>>>>

The ashr package implements a shrinkage-based Empirical Bayes method for
estimating the values of effects $\beta_j$ based on estimates of the effects
($\hat{\beta}_j$) and their standard errors ($s_j$).

The default model is that the effects $\beta_j$ are identically distributed
from a unimodal distribution $g$. In particular, ashr assumes that $\beta_j$
is independent of the standard error $s_j$. That is,

$$\beta_j | s_j \sim g(\cdot) \quad (*).$$

Here we consider the more general model:

$$\beta_j/s_j^\alpha | s_j \sim g(\cdot)   \quad (**).$$

The case $\alpha = 0$ is the model (*) above. The case $\alpha = 1$ assumes
that the *t* statistics $\beta_j/s_j$ are identically distributed from
a unimodal distribution. Under this assumption the expected size of the
*unstandardized* effect $\beta_j$ depends on the standard error $s_j$; the
larger $s_j$ is, the larger (in absolute value) $\beta_j$ is expected to be.

By analogy with Wen and Stephens (AoAS), we refer to the model (*) as the
EE model ("exchangeable effects"), and the model (**) as the ET model
("exchangeable T statistics model").

What might motivate the ET model? We can provide two distinct motivations.
First, suppose for concreteness that the effects $\beta_j$ reflect differences
in gene expression between two conditions. One factor that affects $s_j$ is
the variance of gene $j$ within each condition. One could imagine that, perhaps,
genes with a larger variance within each condition are less tightly regulated,
and therefore more likely to show a large difference between conditions
(*i.e.* large $\beta_j$) than genes with a small variance. This provides
a biological motivation for the possibility that larger $s_j$ might
correlate with larger $\beta_j$ (although of course not for the exact
functional form above).

A second motivation is more statistical: it turns out that this assumption is
in some sense the implicit assumption made by existing methods to fdr analysis
based on p values. More specifically, under this alternative assumption, when
attempting to identify "significant" effects, the empirical Bayes approach
will rank the genes in the same way as the usual $p$ values computed from
$\hat{\beta}_j/s_j$.

It is straightforward to use the ashr package to perform analysis under the
ET model: simply specify alpha = 1. (Internally, this causes ash to replace
betahat with the standardized betahat, $\hat{\beta}_j/s_j$, and the standard
errors for these standarized betahat values with 1; there there is some
bookkeeping to be done to make sure we return the right likelihoods and
posteriors for the original beta, and not for these standardized values...
ash takes care of this.) It is also straightforward to compare
the two competing modelling assumptions (EE vs ET) by computing the
log-likelihood ratio,

$$\log\{p_{EE}(\hat{\beta} | s, \hat{g}_{EE}) / p_{ET}(\hat{\beta} | s, \hat{g}_{ET})\}.$$

We now illustrate by a simulated example. We assume that the standard errors
come from a gamma distribution, and then generate the effects $\beta_j$ under
the ET model so that genes with bigger standard errors tend to have bigger
effects.

```{r generate_data}
set.seed(1234)
nsamp      = 1000
betahat.se = rgamma(nsamp,1,1)
beta       = betahat.se * rnorm(nsamp) #simulate effects under the alternative assumption
betahat    = rnorm(nsamp,beta,betahat.se)
zscore     = betahat/betahat.se
pval       = pchisq(zscore^2,df=1,lower.tail=FALSE)
plot(betahat,-log(pval))
```
